---
title: "Salary Survey Data Cleaning Pipeline Report"
author: "Shoaib"
date: "`r Sys.Date()`"
format:
  docx:
    toc: true
    number-sections: true
    fig-width: 6.5
    fig-height: 4
    fig-dpi: 150
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
source("../scripts/00_config.R")

library(dplyr)
library(ggplot2)
library(stringr)
library(scales)

df_raw     <- readRDS(paste0("../", raw_path, "raw_data.rds"))
df_clean   <- readRDS(paste0("../", data_clean_path, "df_clean.rds"))
df_imputed <- readRDS(paste0("../", data_clean_path, "df_imputed.rds"))
data_dict  <- read.csv(paste0("../", outputs_path, "data_dictionary.csv"))

df_imputed$additional_comp <- as.numeric(df_imputed$additional_comp)

colnames(df_raw) <- c("timestamp", "age", "industry", "job_title", "job_title_context",
                       "annual_salary", "additional_comp", "currency", "currency_other",
                       "income_context", "country", "us_state", "city", "exp_overall",
                       "exp_field", "education", "gender", "race")

report_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    axis.title    = element_text(size = 10),
    axis.text     = element_text(size = 9),
    legend.title  = element_text(size = 10),
    legend.text   = element_text(size = 9),
    plot.margin   = margin(10, 10, 10, 10)
  )

theme_set(report_theme)
```

# Dataset Overview

**Source:** Ask A Manager Salary Survey 2021

**Raw size:** `r nrow(df_raw)` rows and `r ncol(df_raw)` columns

**Download date:** April 2026

**Variables:** 18 variables covering salary, demographics, work experience and location

**Pipeline outputs:** Cleaned dataset exported in four formats - .sav, .dta, .xlsx, .rds

**Assigned scope:** All 18 variables in the dataset

# Data Quality Diagnostics Before Cleaning

## Missingness Overview Before and After Cleaning

```{r missingness-chart}
#| fig-width: 6.5
#| fig-height: 5

pct_before <- round(colSums(is.na(df_raw[, 1:18])) / nrow(df_raw) * 100, 2)
pct_after  <- round(colSums(is.na(df_imputed[, 1:18])) / nrow(df_imputed) * 100, 2)

var_names <- c("timestamp", "age", "industry", "job_title", "job_title_context",
               "annual_salary", "additional_comp", "currency", "currency_other",
               "income_context", "country", "us_state", "city", "exp_overall",
               "exp_field", "education", "gender", "race")

miss_df <- rbind(
  data.frame(variable = var_names, pct_missing = pct_before, stage = "Before cleaning"),
  data.frame(variable = var_names, pct_missing = pct_after,  stage = "After cleaning")
)

miss_df <- miss_df[miss_df$pct_missing > 0, ]

ggplot(miss_df, aes(x = reorder(variable, pct_missing), y = pct_missing, fill = stage)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Before cleaning" = "#2E4057", "After cleaning" = "#4A90A4")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title = "Percentage Missing by Variable Before and After Cleaning",
    x     = NULL,
    y     = "Percent Missing (%)",
    fill  = NULL
  ) +
  theme(legend.position = "bottom")
```

## Key Issues Identified

- **Country:** 390 unique values found - USA alone had over 50 spelling variants
- **Salary:** Range 0 to 6,000,070,000 - 136 entries below 1,000, 34 zeros
- **Additional compensation:** 7,372 missing (26%) - survey wording suggests no bonus
- **Experience logic:** 265 entries where field experience exceeded overall experience
- **City:** 187 remote worker entries, 121 missing, 10 invalid entries
- **Gender:** Two near-identical categories merged into one standard category

# Cleaning Decisions

## Stage 1 - Whitespace Removal
Leading and trailing whitespace stripped from all text columns. 1,708 country entries had invisible spaces causing duplicate values.

## Stage 2 - Timestamp Flagging
618 entries outside the 2021 survey period flagged but retained in the dataset.

## Stage 3 - Country Normalisation
All known country variants mapped to canonical names. 26 invalid entries flagged. USA variants reduced from over 50 to one standard value.

## Stage 4 - Salary Outlier Flagging
IQR method applied. 34 zero salaries, 102 below 1,000 and 1,218 above the upper fence of 192,750 flagged. No rows dropped.

## Stage 5 - Additional Compensation Flagging
7,372 missing and 2,045 above 500,000 flagged.

## Stage 6 - Industry Normalisation
96 lowercase entries corrected to title case. Library and Libraries merged into Library.

## Stage 7 - Job Title Flagging
1 missing and 2 invalid entries flagged.

## Stage 8 - Gender Standardisation
Prefer not to answer (1 entry) merged into Other or prefer not to answer (298 entries).

## Stage 9 - Experience Band Formatting
The band 5-7 years standardised to 5 - 7 years across both experience columns for consistency.

## Stage 10 - Experience Logic Checks
265 impossible field vs overall experience entries flagged. 12 impossible age vs experience combinations flagged.

## Stage 11 - City Flagging
121 missing, 187 remote entries and 10 invalid entries flagged.

## Stage 12 - Multi-State Entries
114 US state entries with multiple states reduced to first state listed.

## Stage 13 - Currency Other Standardisation
All values converted to uppercase. 8 numeric and 10 non-currency entries flagged.

## Stage 14 - Currency Country Cross Check
110 USD entries for clearly non-US countries flagged for transparency.

## Stages 15 and 16 - Missing Education and Race
240 missing education and 196 missing race entries flagged for imputation.

# Imputation Applied

```{r imputation-chart}
#| fig-width: 6.5
#| fig-height: 3.5

imp_df <- data.frame(
  variable = c("additional_comp", "race", "gender", "education"),
  method   = c("Constant", "Constant", "Constant", "Mode"),
  count    = c(7372, 196, 185, 240)
)

ggplot(imp_df, aes(x = reorder(variable, count), y = count, fill = method)) +
  geom_col(width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("Mode" = "#2E4057", "Constant" = "#4A90A4")) +
  scale_y_log10(labels = comma) +
  labs(
    title = "Number of Entries Imputed by Variable and Method (log scale)",
    x     = NULL,
    y     = "Count Imputed (log scale)",
    fill  = "Method"
  ) +
  theme(legend.position = "bottom")
```

| Variable | Method | Value Used | Count | Justification |
|---|---|---|---|---|
| Education | Mode | College degree | 240 | Dominant category at 48% |
| Gender | Constant | Other or prefer not to answer | 185 | Missingness is informative |
| Race | Constant | Prefer not to answer | 196 | Missingness is informative |
| Additional comp | Constant | 0 | 7,372 | Survey wording says if any |

# Post Cleaning Summary

**Rows:** `r nrow(df_imputed)` - no rows were removed during cleaning

**Columns:** 18 original variables retained in the final export

## Final Missingness After Cleaning and Imputation

```{r final-missingness}
final_miss <- data.frame(
  variable      = names(df_imputed)[1:18],
  missing_count = colSums(is.na(df_imputed[, 1:18])),
  pct_missing   = round(colSums(is.na(df_imputed[, 1:18])) / nrow(df_imputed) * 100, 2)
)

final_miss <- final_miss[final_miss$missing_count > 0, ]
knitr::kable(
  final_miss,
  row.names = FALSE,
  col.names = c("Variable", "Missing Count", "Percent Missing"),
  caption   = "Variables with remaining missing values after cleaning and imputation"
)
```

## Annual Salary Distribution After Cleaning

```{r salary-dist}
#| fig-width: 6.5
#| fig-height: 3.5

sal_ok <- df_imputed$annual_salary[df_imputed$salary_flag == "ok"]

ggplot(data.frame(salary = sal_ok), aes(x = salary)) +
  geom_histogram(bins = 40, fill = "#2E4057", color = "white", linewidth = 0.2) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Distribution of Annual Salary After Cleaning",
    subtitle = "Outlier-flagged entries excluded from view",
    x        = "Annual Salary",
    y        = "Count"
  )
```

## Additional Compensation Distribution After Cleaning

```{r addcomp-dist}
#| fig-width: 6.5
#| fig-height: 3.5

comp_ok <- df_imputed %>%
  filter(
    !is.na(additional_comp),
    additional_comp > 0,
    additional_comp <= 500000
  ) %>%
  pull(additional_comp)

ggplot(data.frame(comp = comp_ok), aes(x = comp)) +
  geom_histogram(bins = 40, fill = "#4A90A4", color = "white", linewidth = 0.2) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Distribution of Additional Compensation After Cleaning",
    subtitle = "Non-zero values up to 500,000 shown - outliers excluded",
    x        = "Additional Compensation",
    y        = "Count"
  )
```

## Observed vs Imputed Education

```{r obs-vs-imputed}
#| fig-width: 6.5
#| fig-height: 4

edu_all <- rbind(
  data.frame(
    education = c("College degree", "Master's degree", "Some college",
                  "PhD", "Professional degree", "High School"),
    count     = c(13566, 8897, 2096, 1428, 1329, 659),
    status    = "Before imputation"
  ),
  data.frame(
    education = c("College degree", "Master's degree", "Some college",
                  "PhD", "Professional degree", "High School"),
    count     = c(13806, 8897, 2096, 1428, 1329, 659),
    status    = "After imputation"
  )
)

ggplot(edu_all, aes(x = reorder(education, count), y = count, fill = status)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Before imputation" = "#2E4057",
                                "After imputation"  = "#4A90A4")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Education Distribution Before and After Imputation",
    x     = NULL,
    y     = "Count",
    fill  = NULL
  ) +
  theme(legend.position = "bottom")
```

## Median Salary by Education Level

```{r salary-by-education}
#| fig-width: 6.5
#| fig-height: 3.5

sal_edu <- df_imputed %>%
  filter(salary_flag == "ok", !is.na(education)) %>%
  group_by(education) %>%
  summarise(median_salary = median(annual_salary, na.rm = TRUE), .groups = "drop")

ggplot(sal_edu, aes(x = reorder(education, median_salary), y = median_salary)) +
  geom_col(fill = "#2E4057", width = 0.6) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Median Annual Salary by Education Level",
    subtitle = "Based on salary entries with no outlier flag",
    x        = NULL,
    y        = "Median Salary"
  )
```

# Limitations and Known Risks

1. Country names were normalised using manual mapping. Some ambiguous entries may have been incorrectly mapped or flagged.

2. Salary outliers were flagged using the IQR method but were not removed. Analyses using this dataset should filter flagged rows appropriately.

3. Additional compensation missing values were imputed with zero based on survey wording. This assumption may not hold for all respondents.

4. The dataset is self-reported with no verification of salary figures. Extreme values may reflect genuine high earners or data entry errors.

5. Timestamps outside the 2021 survey period suggest the survey remained open beyond the intended collection window. These entries are retained but flagged.

6. Industry is a freetext field with 1,089 unique values. Standardisation was limited to case correction and merging of obvious near-duplicates.

7. 265 entries where field experience exceeded overall experience were flagged but not corrected as the true values are unknown.
